### 2020-04-12 ThesisJo/2_background.tex
2020-04-12

Any wave, regardless of its nature, is governed by a partial differential equation known as the \emph{wave equation}. Its form for a linear, homogeneous (translation-invariant) and isotropic (rotation-invariant) medium is
%
\begin{align}\label{eq_wave_equation}
\nabla^2 f 
 = \frac{1}{c}\frac{\delta^2 f}{\delta t^2},
\end{align}
%\frac{\delta^2 f}{\delta x^2} + \frac{\delta^2 f}{\delta y^2} + \frac{\delta^2 f}{\delta z^2}
%
where $\nabla^2$ is the Laplace operator, $f$ is field strength and $c$ is wave propagation speed. In complex mediums $c$ may depend on signal strength (non-linear), position (non-homogeneous/refractive), angle (anisotropic), frequency (dispersive) or be non-real (lossy). Deep sea is acoustically non-homogeneous due to varying pressure, temperature and salinity, but has negligible dispersion and attenuation. Biological tissue both non-homogeneous, non-linear and lossy. We approximate both cases with (\ref{eq_wave_equation}), because data inaccuracies are inevitable and good for testing algorithmic robustness.

@@

Consider an acoustic signal transmitted to highlight a surface of interest. Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers, stripped of its carrier frequency by heterodyning and its signal signature by matched filtering (Fig. \ref{2_fig_article_relation}). Image reconstruction coherently combine these receiver outputs to focus at an angle or point at a time. Configuring the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

@@

Let the array have $M$ sensors equally spaced at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length). Data sample $n$ from channel $m$ delayed by $\Delta_m$ is written $x_m[n-\Delta_{m}, \vec{p}_{Am}]\in\mathbb{C}$. The delays (normalized by sampling frequency $f_s$) coarsely phase-aligns signals from source $\udot{S}$,
%
\begin{align}
\Delta_m[n] = \frac{\lVert \vec{p}_{mS}\rVert - \underset{m'}{\mathrm{min}}\ \lVert \vec{p}_{m'S}\rVert}{c\,f_s},\label{2_eq_delta}
\end{align}
%
where both $\vec{p}_{mS}$ and $\vec{p}_{m'S}$ denote position vector $\udot{m}$ to $\udot{S}$. Subtracting the minimum distance is mathematically optional, but facilitates physical realization and low latency. By assuming implicit dependence on $\vec{p}$ and data pre-delayed to the pixel of interest, we simplify notation to $x_m[n]\in\mathbb{C}$. Beamformer output $z[n]\in\mathbb{C}$ is, by definition, the weighted sum of delayed data samples,
%
\begin{align}
z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]} \in\mathbb{C},\label{2_eq_z}
\end{align}
%
where the weights are normalized and $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. Spatial responses of the weights for an array steered to broadside are characterized by the beam pattern,
%
\begin{align}
W(\vec{k}) = \sum_m^{M-1} w_m^* e^{j\vec k \cdot \vec{p}_{Am}},\label{2_eq_beam pattern}
\end{align}
%. Beampattern is broadside-steered beamformer output in the presence of a single plane wave, as a function of the wave's angle.
%%Steered response is beamformer's output in the presence of some wave field, as a function of steering angle. 
%
where the start position $\udot{A}$ is arbitrary. Note how the beam pattern depends on both angle and frequency of the incoming wave.

@@

To make MVDR robust, it must handle avoid signal cancellation, preserve speckle statistics and ensure numerical and statistical robustness. The literature suggests estimating $\R$ with a sample covariance matrix $\eR$ using some degree of
%
\begin{itemize}
\item \emph{Spatial smoothing}: Decorrelates coherent echoes to avoid signal cancellation, assuming regular arrays. We achieve this through subarray averaging~\cite{Kailath1985}, which trades off spatial resolution and makes the covariance matrix exhibit Toeplitz-like structure. Some studies construct Toeplitz-matrices directly since are simpler to invert [TK87]. $\eR$ is guaranteed non-singular in cases with twice as many subarrays as signals [WPMS88,PK89]. 
\item \emph{Temporal smoothing} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009};
\item \emph{Diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
\end{itemize}%
%
These techniques ensure a numerically well conditioned $\eR$, allow reducing the computational complexity and expose enough data parallelism to facilitate GPU-acceleration.

@@

In imaging sonar and ultrasound systems, the combined transmit and receive response is usually narrow enough to contain most energy in three or five center beams~\cite{Nilsen2009a,VanTrees2002}. This allows  Above 32 channels the inversion step becomes a greater computational bottleneck than covariance estimation. We mitigate this by applying space reduction in the spatial frequency-domain, or beamspace.

@@

% \subsection{Image quality}

>>>>

Any wave, regardless of its nature, is governed by a partial differential equation known as the \emph{wave equation}. Its form for a linear, homogeneous (translation-invariant) and isotropic (rotation-invariant) medium is
%
\begin{align}\label{eq_wave_equation}
\nabla^2 f 
 = \frac{1}{c}\frac{\delta^2 f}{\delta t^2},
\end{align}
%\frac{\delta^2 f}{\delta x^2} + \frac{\delta^2 f}{\delta y^2} + \frac{\delta^2 f}{\delta z^2}
%
where $\nabla^2$ is the Laplace operator, $f$ is field strength and $c$ is wave propagation speed. In complex mediums $c$ may depend on signal strength (non-linear), position (non-homogeneous/refractive), angle (anisotropic), frequency (dispersive) or be non-real (lossy). Deep sea is acoustically non-homogeneous due to varying pressure, temperature and salinity, but has negligible dispersion and attenuation. Biological tissue are non-homogeneous, non-linear and lossy. We approximate both cases with (\ref{eq_wave_equation}), because data inaccuracies are inevitable and good for testing algorithmic robustness.

@@

Consider an acoustic signal transmitted to highlight a surface of interest. Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers, stripped of its carrier frequency by heterodyning and its signal signature by matched filtering (Fig. \ref{2_fig_article_relation}). Further assume the sensors are linear, infinite bandwidth and invarant to shifts in position (homogeneous, inter-independent), direction (isotropic) and time. Image reconstruction coherently combines these receiver outputs to focus at an angle or point at a time. Configuring the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

@@

Let the array have $M$ sensors equally spaced at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length). Data sample $n$ from channel $m$ delayed by $\Delta_m$ is written $x_m[n-\Delta_{m}, \vec{p}_{Am}]\in\mathbb{C}$. The delays (normalized by sampling frequency $f_s$) coarsely aligns signals from source $\udot{S}$,
%
\begin{align}
\Delta_m[n] = \frac{\lVert \vec{p}_{mS}\rVert - \underset{m'}{\mathrm{min}}\ \lVert \vec{p}_{m'S}\rVert}{c\,f_s},\label{2_eq_delta}
\end{align}
%
where both $\vec{p}_{mS}$ and $\vec{p}_{m'S}$ denote position vector $\udot{m}$ to $\udot{S}$. Subtracting the minimum distance is mathematically optional, but facilitates physical realization and low latency. By assuming implicit dependence on $\vec{p}$ and data pre-delayed to the pixel of interest, we simplify notation to $x_m[n]\in\mathbb{C}$. Beamformer output $z[n]\in\mathbb{C}$ is, by definition, the weighted sum of delayed data samples,
%
\begin{align}
z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]} \in\mathbb{C},\label{2_eq_z}
\end{align}
%
where the weights are normalized and $w_m=\in\mathbb{C}$ is the weight factor assigned to channel $m$. The weights are normalized, $\dvec{w}^H\dvec{1}=1$, Spatial responses of the weights for an array steered to broadside are characterized by the beam pattern,
%
\begin{align}
W(\vec{k}) = \sum_m^{M-1} w_m^* e^{j\vec k \cdot \vec{p}_{Am}},\label{2_eq_beam pattern}
\end{align}
%. Beampattern is broadside-steered beamformer output in the presence of a single plane wave, as a function of the wave's angle.
%%Steered response is beamformer's output in the presence of some wave field, as a function of steering angle. 
%
where the start position $\udot{A}$ is arbitrary. Note how the beam pattern depends on both angle and frequency of the incoming wave.

@@

To make MVDR robust, it must avoid signal cancellation, preserve speckle statistics and ensure numerical and statistical robustness. Literature suggests estimating $\R$ with a sample covariance matrix $\eR$ using some degree of
%
\begin{itemize}
\item \emph{Spatial smoothing}: Decorrelates coherent echoes to avoid signal cancellation, assuming regular arrays. We achieve this through subarray averaging~\cite{Kailath1985}, which trades off spatial resolution and makes the covariance matrix exhibit Toeplitz-like structure. Some studies construct inversion-friendly Toeplitz-matrices directly [TK87]. $\eR$ is guaranteed non-singular in cases with twice as many subarrays as signals [WPMS88,PK89]. 
\item \emph{Temporal smoothing} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009};
\item \emph{Diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
\end{itemize}%
%
These techniques ensure a numerically well conditioned $\eR$, allow reducing the computational complexity and expose enough data parallelism to facilitate GPU-acceleration.

@@

In imaging sonar and ultrasound systems, the combined transmit and receive response is usually narrow enough to contain most energy in three or five center beams~\cite{Nilsen2009a,VanTrees2002}. Above 32 channels the inversion step becomes a greater computational bottleneck than covariance estimation. We mitigate this by applying space reduction in the spatial frequency-domain, or beamspace.

@@

\section{Image synthesis}

\subsection{Template matching}

Template matching classifies objects by comparing them to idealized image templates of relevant objects and orientations, and assigning the most similar class. Similarity metrics like sum of absolute differences (SMD) and normalized cross-correlation (NCC) yield good results in computer vision~\cite{Brunelli2009}, but not in sonar imaging---likely due to noise~\cite{Myers2007}.

To improve noise- In sonar imaging Several adaptiations to sonar imaging approaches

### 2020-04-12 ThesisJo/3_publications.tex
2020-04-12

We find that implementing MVDR on GPUs instead of CPUs accelerates it by two orders of magnitude, but requires considerable and delicate efforts to reduce arithmetic operations and fine-tune cache usage. The design is limited to 32 channels due to data dependencies that are complicated and not inherently parallel, combined with the relatively limited bandwidth of this hardware architecture.

>>>>

We find that implementing MVDR on GPUs instead of CPUs accelerates it by two orders of magnitude, but requires considerable and delicate efforts to reduce arithmetic operations and fine-tune cache usage.

Processing of pixels are independent of one another, but processing
At the pixel level Data-parallelism Pixels are processed independently,

The design is limited to 32 channels, due to data dependencies that are complicated and not inherently parallel at the sub-pixel level, combined with the relatively limited bandwidth of this hardware architecture.

### 2020-04-12 ThesisJo/frontmatter.tex
2019-11-02
2020-04-12

>>>>

%\newpage
\chapter*{Symbol reference}
\addcontentsline{toc}{chapter}{Symbol reference}

%\begin{table}[H]\noindent%
%\begin{tabular}{>{$}l<{$} >{}l<{} l >{$}l<{$}}
\newcommand\pb[1]{\parbox[c]{9mm}{\textbf{\centering#1}}}
\newcommand\bb[1]{\mathbb{#1}}
%@{\hspace*{0\tabcolsep}}
\begin{longtable}{@{\hspace*{-\tabcolsep}}>{$}l<{$}   >{$\mathbb}l<{$}  l  >{$}l<{$}  @{\hspace*{-\tabcolsep}}>{\footnotesize}l }
\hspace*{-.5cm}\textbf{Symbol} & \textbf{Space} & \textbf{Description}                                              & \textbf{Equal to}                     \\\endhead
	\mathcal{A}          &                 & Name of geometrical object (e.g. an \textbf{A}rray)                     &                                       \\
	\ubar{A}             & \bb V^3\bb A    & Frame (configuration) of $\mathcal{A}$                                  & (\uvec{A},\udot{A})                   \\
	\uvec{A}             & \bb V^3         & Orientation of $\mathcal{A}$ (absolute, coordinate free)                &                                       \\
	\udot{A}             & \bb A           & Position of $\mathcal{A}$ (absolute, coordinate free)                   &                                       \\
	\hat{a}              & \bb V^3         & Steering vector (unit-length, coordinate-free)                          &                                       \\
	\dvec{a}             & \bb R^3         & Steering vector (unit-length, decomposed)                               &                                       \\
	\dvec{B}             & \bb C^{N_b,L}   & Butler/beam-space matrix (sparse DFT matrix)                            &                                       \\
	c                    & \bb R^+         & Wave propagation speed [m/s]                                            &                                       \\
	\Delta_m             & \bb R           & Delay applied to $m$th sensor (normalized by $f_s$)                     &                                       \\
	\dvec{\Delta}        & \bb R^M         & Delay vector, all sensors (normalized by $f_s$)                         &                                       \\
	f                    & \bb R           & Field strength in (\ref{eq_wave_equation})                              &                                       \\
	                     & \bb R^+         & Frequency (everywhere else) [Hz]                                        &                                       \\
   f_s                  & \bb R^+         & Temporal sampling frequency [Hz]                                        &                                       \\
	F                    & \bb R^+         & Temporal sampling frequency (normalized by $f_s$)                       & \frac{f}{f_s}                         \\
	\vec{k}              & \bb V^3         & Wave vector, wavenumber with direction                                  & \lVert k\rVert \hat{a} &(1)           \\
	\lVert\vec{k}\rVert  & \bb R^+         & Spatial angular frequency (wavenumber) [1/m]                            & \frac{2\pi}{\lambda},\frac{\omega}{c} \\
	K                    & \bb Z           & Number of temporal samples to average (one-sided)                       &                                       \\
	\lambda              & \bb R^+         & Wavelength [m]                                                          &                                       \\
	L                    & \bb N           & Number of elements in subarray                                          &                                       \\
	M                    & \bb N           & Number of sensors (hydrophones/microphones)                             &                                       \\
	n                    & \bb Z           & Temporal sample index                                                   & f_s t                                 \\
	N                    & \bb N           & Number of temporal samples                                              &                                       \\
	N_K                  & \bb N           & Number of temporal samples to average                                   & 2K+1                                  \\
	N_L                  & \bb N           & Number of subarrays                                                     & M-L+1                                 \\
	\vec{p}_{AB}         & \bb V^3         & Position vector (coordinate-free) from $\udot{A}$ to $\dot{B}$          & \udot{B}-\dot{A}                      \\
	\dvec{p}_{AB}^C      & \bb R^3         & Position vector from $\udot{A}$ to $\dot{B}$, represented in $\uvec{C}$ &                                       \\
	\phi                 & \bb R           & Azimuth steering angle                                                  &                                       \\
	\varphi              & \bb R           & Elevation angle                                                         &                                       \\
	\dvec{R}             & \bb C^{M,M}     & Spatial covariance matrix                                               &                                       \\
	\R_{\mathrm{BS}}     & \bb C^{M',M'}   & Spatial covariance matrix (beam-space)                                  &                                       \\
	\eR                  & \bb C^{L,L}     & Sample spatial covariance matrix                                        &                                       \\
	\eR_{\mathrm{BS}}    & \bb C^{N_b,N_b} & Sample spatial covariance matrix (beam-space)                           &                                       \\
	t                    & \bb R           & Time [s]                                                                & \frac{n}{f_s}                         \\
	T_s                  & \bb R           & Temporal sampling period [s]                                            & F_s^{-1}                              \\
	\theta               & \bb R           & Azimuth angle                                                           & \omega t - \vec k\vec p, \Omega n                    \\
	\omega               & \bb R^+         & Temporal angular frequency [1/s]                                        & 2\pi f                                \\
   \Omega               & \bb R^+         & Temporal angular frequency (normalized by $f_s$)                        & 2\pi F                                \\
	w_m                  & \bb C^N         & Weight on $m$th sensor                                                  &                                       \\
	\dvec{w}             & \bb C^{M,N}     & Weight set, all sensors                                                 &                                       \\
	x_m                  & \bb C^N         & Data (delayed, element-space) from $m$th sensor                         &                                       \\
	x_{\textrm{BS},m}    & \bb C^{N_b}     & Data (delayed, beam-space) from $m$th sensor                            &                                       \\
	\dvec{x}             & \bb C^{M,N}     & Data (delayed, element-space) from all sensors                          &                                       \\
	\dvec{x}_\mathrm{BS} & \bb C^{M,N_b}   & Data (delayed, beam-space) from all sensors                             &                                       \\
	z                    & \bb C^N         & Beamformer output                                                       &                                       \\
\multicolumn{4}{l}{{\footnotesize (1)}\small\textit\quad{}Assumes linear, homogeneous medium.}
\end{longtable}
%	y_m                  & \bb C^N         & Data (no delay, element-space) from $m$th sensor                        &                                       \\
%\end{tabular}
%\end{table}

%\newpage
\chapter*{Acronym reference}
\addcontentsline{toc}{chapter}{Acronym reference}

\begin{longtable}{l l}
	\textbf{Acronym} & \textbf{Description}                                                      \\
	ADC              & Analog-to-digital converter                                               \\
	ATR              & Automatic target recognition                                              \\
	AUV              & Autonomous underwater vehicle                                             \\
	BS-Capon         & Beam-space Capon (or MVDR, beamformer)                                    \\
	CPU              & Central processing unit                                                   \\
	CUDA             & Compute unified device architecture (from Nvidia)                         \\
	DAC              & Digital-to-analog converter                                               \\
	DAS              & Delay-and-sum (beamformer)                                                \\
	DFT              & Discrete Fourier transform                                                \\
	DMA              & Direct memory access                                                      \\
	ES-Capon         & Element-space Capon (or MVDR, beamformer)                                 \\
	FFT              & Fast Fourier transform                                                    \\
	FPGA             & Field-programmable gate array                                             \\
	FPR              & False positive rate                                                       \\
	FWHM             & Full width at half maximum                                                \\
	GPU              & Graphics processing unit                                                  \\
	HISAS1030        & High resolution synthetic aperture sonar, Kongsberg Maritime (model 1030) \\
	HPC              & High performance computing                                                \\
	HUGIN            & An autonomous underwater vehicle developed by Kongsberg Maritime          \\
	LCA              & Low complexity adaptive (beamformer)                                      \\
   MCM & Mine counter measures \\
	MKL              & Math kernel library (from Intel)                                          \\
	MVDR             & Minimum variance distortionless response (or Capon, beamformer)           \\
	NDC              & Normalized device coordinates                                             \\
	PCI (Express)    & Peripheral component interconnect (data bus)                              \\
	RAM              & Random access memory                                                      \\
	ROC              & Receiver operating curve                                                  \\
	SAS              & Synthetic aperture sonar                                                  \\
	SDFT             & Sliding discrete Fourier transform                                        \\
	SIGMAS+          & Name of an acoustic GPU-based simulator                                   \\
	SSS              & Sidescan sonar                                                            \\
	SM               & Streaming multi-processor \\
   TPR & True positive rate 
\end{longtable}
